---
title: "Reproducibility Report: Maglio & Polman (2014), Study 1"
subtitle: "Bumper-rails version"
author: "Sarah Segall"
date: today
format:
  pdf:
    toc: true
    toc-location: left
    embed-resources: true
execute:
  error: true
---

:::{.callout-note title="Goals"}
1. Load a published dataset with `here()` and `read_excel()`
2. Reproduce the ANOVA and effect sizes reported in the paper
3. Say clearly what did and did not reproduce, and why
:::

This is the **bumper-rails** version: most of the code is written for you, and
you replace each `"FIXME"` with the right variable name or expression. If you
would rather write the code yourself, use
`reproducibility-report-minimal.qmd` instead. Both reproduce the same results.

`error: true` is set at the top of this file, so the document still renders
while it has `FIXME`s in it. The chunks you have not finished yet show their
error message in red, which is a handy to-do list.

Before you start, read Study 1 in the paper:
`original_paper/maglio-polman-2014.pdf` (two pages).

## The study

Maglio, S. J., & Polman, E. (2014). Spatial orientation shrinks and expands
psychological distance. *Psychological Science, 25*(7), 1345–1352.
<https://doi.org/10.1177/0956797614530571>

**Methods summary.** Researchers recruited 202 volunteers at the Bay Street
subway station in Toronto. Half were waiting on the westbound platform and half
on the eastbound platform. Each participant was randomly assigned to rate how far
away one of four stations felt (1 = very close, 7 = very far): Spadina (two stops
west), St. George (one stop west), Bloor-Yonge (one stop east), or Sherbourne
(two stops east). The authors ran a 2 × 4 ANOVA on the distance ratings, then
compared eastbound and westbound travelers separately at each station.

**Target outcomes.** This is the result you are trying to reproduce, quoted from
the Study 1 results section:

> We carried out a 2 (orientation: toward, away from) × 4 (station: Spadina, St.
> George, Bloor-Yonge, Sherbourne) analysis of variance (ANOVA) on closeness
> ratings, which revealed no main effect of orientation, F < 1, and a main effect
> of station, F(3, 194) = 24.10, p < .001, ηp² = .27. This main effect was
> qualified by the predicted interaction between orientation and station, F(3,
> 194) = 16.28, p < .001, ηp² = .20. We decomposed this interaction by the
> subjective-distance ratings between participants traveling east and west for
> each of the four subway stations. Westbound participants rated the stations to
> the west of Bay Street as closer than did eastbound participants; this effect
> was obtained for both the station one stop to the west (St. George, p < .001,
> ηp² = .28) and the station two stops to the west (Spadina, p = .001, ηp² =
> .20). The opposite pattern held true for stations to the east of Bay Street.
> Eastbound participants rated the stations to the east of Bay Street as closer
> than did westbound participants; this effect was obtained for both the station
> one stop to the east (Bloor-Yonge, p = .053, ηp² = .08) and the station two
> stops to the east (Sherbourne, p < .001, ηp² = .24). Figure 1 summarizes these
> results.

ηp² is **partial eta squared**, an effect size. You will compute it below.

## Step 1: Load packages

`readxl` (reads Excel files) and `broom` (turns model output into tibbles) are
installed along with the tidyverse, but `library(tidyverse)` does not load them,
so we load them by name.

```{r}
#| eval: false
install.packages(c("tidyverse", "here"))
```

```{r}
#| message: false
library(tidyverse)
library(readxl)
library(broom)
library(here)
```

## Step 2: Load the data

The authors posted one Excel file per study. We only need Study 1. The columns
are described in `docs/codebook.md`.

```{r}
d <- read_excel(here("data", "S1_Subway.xlsx"))

glimpse(d)
```

## Step 3: Tidy the data

The data are already tidy: one row per participant, one column per variable.
But R has read `DIRECTION` and `STN_NAME` in as text. For an ANOVA they need to
be **factors**, and we set the order of the station levels to run west to east,
so tables and plots come out in map order.

```{r}
d <- d |>
  mutate(
    DIRECTION = factor(DIRECTION),
    STN_NAME = factor(STN_NAME, levels = c("SPAD", "STG", "B-Y", "SHER"))
  )

glimpse(d)
```

## Step 4: Reproduce the analysis

### How many people are in each cell?

Count the participants in each direction × station combination. Replace each
`"FIXME"` with a column name.

```{r}
d |>
  count(DIRECTION, STN_NAME)
```

Does this match the sample size in the paper (202 total, 101 per platform)?
Is the design balanced, with the same *n* in every cell?

### The 2 × 4 ANOVA

When a paper says "analysis of variance", a good first step in R is `aov()`.
The formula `y ~ A * B` asks for both main effects *and* their interaction; it
is shorthand for `y ~ A + B + A:B`.

Fill in the formula with the outcome (the distance rating) and the two
predictors.

```{r}
mod <- aov(DISTANCE ~ DIRECTION * STN_NAME, data = d)

summary(mod)
```

The numbers in parentheses in "F(3, 194)" are the degrees of freedom: 3 for the
effect, 194 for the residuals. Check that yours match.

The teaching team gets:

- Main effect of direction: F(1, 194) = 0.66
- Main effect of station: F(3, 194) = 23.35
- Interaction between direction and station: F(3, 194) = 16.28

Two of those match the paper. One is close but not the same. Keep going; the
bonus question at the end asks you to work out why.

### Partial eta squared

Partial eta squared is the share of variance an effect accounts for, once
the other effects are set aside:

$$
\eta_p^2 = \frac{SS_\text{effect}}{SS_\text{effect} + SS_\text{residual}}
$$

`tidy()` from broom turns the ANOVA table into a tibble, with one row per term
and the sums of squares in the `sumsq` column. That means we can compute
ηp² with ordinary tidyverse verbs.

```{r}
anova_table <- tidy(mod)
anova_table
```

```{r}
ss_residual <- anova_table |>
  filter(term == "Residuals") |>
  pull(sumsq)

anova_table |>
  filter(term != "Residuals") |>
  mutate(partial_eta_sq = sumsq / (sumsq + ss_residual)) |>
  select(term, df, statistic, p.value, partial_eta_sq)

```

Do your ηp² values round to the paper's .27 (station) and .20 (interaction)?

### Station-by-station comparisons

Next the authors compare eastbound and westbound travelers at each station
separately. A one-way ANOVA with a single two-level predictor is the same test
as a *t*-test (F = t²). With only one effect in the model, ηp² is just
SS_direction / SS_total.

#### St. George

```{r}
#| label: st-george

# 1. Keep only the St. George ratings
st_george <- d |>
  filter(STN_NAME == "STG")

# 2. Which group rated St. George as farther away, eastbound or westbound?
#    No statistics yet, just the mean rating in each direction.
st_george |>
  group_by(DIRECTION) |>
  summarize(mean_distance = mean(DISTANCE), n = n())

# 3. One-way ANOVA: distance predicted by direction
st_george_mod <- aov(DISTANCE ~ DIRECTION, data = st_george)

st_george_table <- tidy(st_george_mod)
st_george_table

# 4. Partial eta squared (reuse your code from above)
st_george_residual <- st_george_table |>
  filter(term == "Residuals") |>
  pull(sumsq)

st_george_table |>
  filter(term != "Residuals") |>
  mutate(partial_eta_sq = sumsq / (sumsq + st_george_residual)) |>
  select(term, df, statistic, p.value, partial_eta_sq)

```

#### Spadina, Bloor-Yonge, and Sherbourne

Repeat the four steps for the other three stations. Copying the St. George
chunk three times works. If you want to try something tidier, the
**Going further** section below does all four stations in one pipeline.

```{r}
#| label: spadina

# 1. Keep only the spadina ratings
spadina <- d |>
  filter(STN_NAME == "SPAD")

# 2. Which group rated St. George as farther away, eastbound or westbound?
#    No statistics yet, just the mean rating in each direction.
spadina |>
  group_by(DIRECTION) |>
  summarize(mean_distance = mean(DISTANCE), n = n())

# 3. One-way ANOVA: distance predicted by direction
spadina_mod <- aov(DISTANCE ~ DIRECTION, data = spadina)

spadina_table <- tidy(spadina_mod)
spadina_table

# 4. Partial eta squared (reuse your code from above)
spadina_residual <- spadina_table |>
  filter(term == "Residuals") |>
  pull(sumsq)

spadina_table |>
  filter(term != "Residuals") |>
  mutate(partial_eta_sq = sumsq / (sumsq + spadina_residual)) |>
  select(term, df, statistic, p.value, partial_eta_sq)

```

```{r}
#| label: bloor-yonge

bloor_yonge <- d |>
  filter(STN_NAME == "B-Y")

bloor_yonge |>
  group_by(DIRECTION) |>
  summarize(mean_distance = mean(DISTANCE), n = n())

bloor_yonge_mod <- aov(DISTANCE ~ DIRECTION, data = bloor_yonge)

bloor_yonge_table <- tidy(bloor_yonge_mod)
bloor_yonge_table

bloor_yonge_residual <- bloor_yonge_table |>
  filter(term == "Residuals") |>
  pull(sumsq)

bloor_yonge_table |>
  filter(term != "Residuals") |>
  mutate(partial_eta_sq = sumsq / (sumsq + bloor_yonge_residual)) |>
  select(term, df, statistic, p.value, partial_eta_sq)

```

```{r}
#| label: sherbourne

sherbourne <- d |>
  filter(STN_NAME == "SHER")

sherbourne |>
  group_by(DIRECTION) |>
  summarize(mean_distance = mean(DISTANCE), n = n())

sherbourne_mod <- aov(DISTANCE ~ DIRECTION, data = sherbourne)

sherbourne_table <- tidy(sherbourne_mod)
sherbourne_table

sherbourne_residual <- sherbourne_table |>
  filter(term == "Residuals") |>
  pull(sumsq)

sherbourne_table |>
  filter(term != "Residuals") |>
  mutate(partial_eta_sq = sumsq / (sumsq + sherbourne_residual)) |>
  select(term, df, statistic, p.value, partial_eta_sq)


```

Fill in this table as you go:

| Station | Paper *p* | Your *p* | Paper ηp² | Your ηp² |
|---|---|---|---|---|
| St. George | < .001 | < .001 | .28 | .28 |
| Spadina | .001 | .001 | .20 | .20 |
| Bloor-Yonge | .053 | .053 | .08 | .08 |
| Sherbourne | < .001 | < .001 | .24 | .24 |


## Step 5: Reflection

Were you able to reproduce the results you attempted to reproduce? If not, which
parts could you not reproduce?

> I was able to replicate the interaction effect and find no main effect of orientation. I was also able to reproduce the p-values and partial theta squared vlaues for all four station-by-station comparisons. I could *not* reproduce the F value for the main effect of station - I got F(3, 194) = 23.35 and the paper reports 24.10.

How difficult was it to reproduce the results?

> It was not bad because the dataset was organized. The whole analysis is one ANOVA and four follow-up tests. 

What made it hard, and what made it easy? Think about the data files, the
variable names, and how much the paper told you about the analysis.

> It was easy in that the authors posted the raw data and had their data organized by study. The paper also reported all of the relevant statistics which was good. It was difficult in that the paper did not mention the software they used. It was also a little confusing that the paper called the factor orientation rather than direction and that the station names were all abbreviated.

## Bonus questions

**1. Why is the station F different?** Your F for station (23.35) is close to
the paper's (24.10) but not equal, while the interaction matches exactly. Try
refitting the model with the predictors in the opposite order,
`DISTANCE ~ STN_NAME * DIRECTION`. Does anything change? Look back at your cell
counts. What does that suggest about how the authors' software (SPSS) and R's
`aov()` each divide up the variance? `docs/anova-in-r.md` explains it, but try
first.

```{r}

```

**2. Orientation or direction?** The paper describes the factor as
"orientation: toward, away from". The dataset has `DIRECTION`: east or west.
Are those the same variable? If not, which one did the authors actually put in
their ANOVA? Make an `orientation` column with `if_else()` and refit the model
with it to check.

```{r}

```

**3. Reproduce Figure 1.** Plot the mean distance rating for each station,
split by direction, with error bars of ±1 standard error, as in the paper.
`stat_summary(fun.data = mean_se)` will draw both the means and the error bars
for you.

```{r}

```

## Going further

Splitting, modeling, and tidying once per station can all happen in one
pipeline. `nest()` packs each station's rows into its own small data frame, and
`map()` runs a function on each one:

```{r}
#| eval: false
d |>
  nest(.by = STN_NAME) |>
  mutate(model = map(data, \(df) aov(DISTANCE ~ DIRECTION, data = df)),
         results = map(model, tidy)) |>
  unnest(results)
```

Once that runs, add the ηp² column.
